---
title: "Demultiplexing scMRD data in R"
author: "TR"
date: "2023-04-28"
output: html_document
---

This is a representative example of how we demultiplexed scMRD data in our pre-print here: https://www.biorxiv.org/content/10.1101/2022.09.20.508786v1. The code included here is an updated version of the analysis (may generate figures slightly different than in the pre-print version), and we hope to soon include a link to a peer-reviewed version of our manuscript.  Please note that Our goal was to partition out cells corresponding to each sample without leveraging pathogenic mutation information such that we could then discover mutations in each cell after sample assignment. To do this, we used K-means clustering on germline SNP allele frequency to partition cells into clusters and classify cells based on SNP profile matching.  There are likely alternative ways to approach this, and I am happy to discuss the methods detailed herein as well as alternative approaches. Please direct any code related questions/comments to robinst1@sloankettering.edu. All data needed to reproduce the analyses for this study are deposited in dbGAP in the form of .h5 files and are accessible via study accession phs003233.v1.p1.

We start the analysis by first loading in the required packages and linking to the .h5 file, which is exported off of Mission Bio's Tapestri Pipeline. In this example we'll work with scDNA+Protein run MRD5 from the paper. This code will similarly work with runs MRD2, MRD3, MRD4, and MRD6. The code to demultiplex MRD1 differs as the experimental set up was slightly modified; please refer to the file "scMRD_demultiplexing_MRD1". 
```{r, message=FALSE}

library(tidyverse)
#library(scDNA)
library(BSgenome.Hsapiens.UCSC.hg19)
library(umap)
library(pals)
library(cluster)
library(ClusterR)
library(pheatmap)
library(superheat)
library(cultevo)
library(cowplot)
library(lsa)
library(factoextra)
library(NbClust)
library(VariantAnnotation)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(tidyverse)
library(pdist)
library(magrittr)
library(RColorBrewer)
library(ggpubr)
library(BSgenome.Hsapiens.UCSC.hg19)
library(SNPlocs.Hsapiens.dbSNP144.GRCh37)
library(biomaRt)

file <- c("/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX_scDNA_h5_files/WX_MRD5.dna+protein.h5")

```

Functions needed to extract data from the .h5 file
```{r}

read_tapestri_h5_NGT <- function (file, VAF_cutoff = 0.005, GT_cutoff = 20) 
                        {
                            NGT <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/NGT")
                            NGT[NGT == 3] <- NA
                            VAF_select <- which(apply(NGT, MARGIN = 1, function(x) {
                                (sum(!is.na(x))/length(x)) * 100 >= GT_cutoff & (sum(x, 
                                    na.rm = TRUE)/(sum(!is.na(x)) * 2)) >= VAF_cutoff
                            }))
                            NGTlim <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/NGT", 
                                index = list(VAF_select, NULL))
                            NGTlim[NGTlim == 3] <- NA
                            rownames(NGTlim) <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/id", 
                                index = list(VAF_select))
                            colnames(NGTlim) <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ra/barcode")
                            return(NGTlim)
}


quality_filter_NGT <- function (file, NGT, DP_cut = 10, AF_cut = 20, GQ_cut = 20) 
                      {
                          variants <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/id")
                          select_variants <- match(rownames(NGT), variants)
                          AF <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/AF", 
                              index = list(select_variants, NULL))
                          DP <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/DP", 
                              index = list(select_variants, NULL))
                          GQ <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/GQ", 
                              index = list(select_variants, NULL))
                          NGT_filter <- data.frame(setNames(data.frame(AF, variants = rownames(NGT)), 
                              c(tidyselect::all_of(colnames(NGT)), "variants")) %>% 
                              tidyr::pivot_longer(cols = !c(variants), names_to = "Cell", 
                                  values_to = "AF"), data.frame(DP) %>% tidyr::pivot_longer(cols = tidyselect::everything(), 
                              names_to = "Cell", values_to = "DP") %>% dplyr::select(DP), 
                              data.frame(GQ) %>% tidyr::pivot_longer(cols = tidyselect::everything(), 
                                  names_to = "Cell", values_to = "GQ") %>% dplyr::select(GQ), 
                              data.frame(NGT) %>% tidyr::pivot_longer(cols = tidyselect::everything(), 
                                  names_to = "Cell", values_to = "NGT") %>% dplyr::select(NGT)) %>% 
                              dplyr::filter(DP >= DP_cut & GQ >= GQ_cut) %>% dplyr::mutate(pass = dplyr::case_when(NGT == 
                              1 & (AF > AF_cut) & (AF < (100 - AF_cut)) ~ "include", 
                              NGT == 1 & ((AF <= AF_cut) | (AF >= (100 - AF_cut))) ~ 
                                  "exclude", NGT == 2 & AF >= (100 - AF_cut) ~ "include", 
                              NGT == 2 & AF < (100 - AF_cut) ~ "exclude", NGT == 0 & 
                                  AF <= AF_cut ~ "include", NGT == 0 & AF > AF_cut ~ 
                                  "exclude", TRUE ~ "other")) %>% dplyr::filter(.data$pass == 
                              "include") %>% tidyr::pivot_wider(id_cols = .data$Cell, 
                              names_from = .data$variants, values_from = .data$NGT)
                          return(NGT_filter)
                      }

read_tapestri_h5_AF <- function (file)
                         {
                             AF <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/AF")
                             rownames(AF) <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/id")
                             colnames(AF) <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ra/barcode")
                            return(AF)
                        }

quality_filter_AF <- function (file, AF, DP_cut = 10, AF_cut = 20, GQ_cut = 20)
                      {
                          variants <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/id")
                          select_variants <- match(rownames(AF), variants)
                          NGT <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/NGT",
                              index = list(select_variants, NULL))
                          DP <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/DP",
                              index = list(select_variants, NULL))
                          GQ <- rhdf5::h5read(file = file, name = "/assays/dna_variants/layers/GQ",
                              index = list(select_variants, NULL))
                          NGT_filter <- data.frame(setNames(data.frame(NGT, variants = rownames(AF)),
                              c(tidyselect::all_of(colnames(AF)), "variants")) %>%
                              tidyr::pivot_longer(cols = !c(variants), names_to = "Cell",
                                  values_to = "NGT"), data.frame(DP) %>% tidyr::pivot_longer(cols = tidyselect::everything(),
                              names_to = "Cell", values_to = "DP") %>% dplyr::select(DP),
                              data.frame(GQ) %>% tidyr::pivot_longer(cols = tidyselect::everything(),
                                  names_to = "Cell", values_to = "GQ") %>% dplyr::select(GQ),
                              data.frame(AF) %>% tidyr::pivot_longer(cols = tidyselect::everything(),
                                  names_to = "Cell", values_to = "AF") %>% dplyr::select(AF)) %>%
                              dplyr::filter(DP >= DP_cut & GQ >= GQ_cut) %>% dplyr::mutate(pass = dplyr::case_when(NGT ==
                              1 & (AF > AF_cut) & (AF < (100 - AF_cut)) ~ "include",
                              NGT == 1 & ((AF <= AF_cut) | (AF >= (100 - AF_cut))) ~
                                  "exclude", NGT == 2 & AF >= (100 - AF_cut) ~ "include",
                              NGT == 2 & AF < (100 - AF_cut) ~ "exclude", NGT == 0 &
                                  AF <= AF_cut ~ "include", NGT == 0 & AF > AF_cut ~
                                  "exclude", TRUE ~ "other")) %>%
                              dplyr::filter(.data$pass == "include") %>%
                              tidyr::pivot_wider(id_cols = .data$Cell,
                              names_from = .data$variants, values_from = .data$AF)
                          return(NGT_filter)
                     }


```


Load in the manually curated transcript ID table and annotation key. More detailed information for generating the transcript ID table can be found here: https://bowmanr.github.io/scDNA_myeloid/hdf5-and-loom-input.html (this is also overall an excellent resource for scDNA+protein analysis). 
```{r, message=FALSE}

annotation_key <-read.csv("/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/MBio_scDNA_keys/annotation_key_RUNX1C_geneid.csv")
annotation_key$GENEID<- NULL

hg19refseq_txdb <-loadDb("/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/MBio_scDNA_keys/hg19refseq_txdb_new.sqlite")

annotation_key%<>%inner_join(VariantAnnotation::select(hg19refseq_txdb,
                                    keys=annotation_key$ccds_id,
                                    columns=c("TXID","TXNAME"),
                                    keytype = "TXNAME"),
                                    by=c("ccds_id"="TXNAME"))%>%
                                    mutate(TXID=as.character(TXID))

```

Read in the numerical genotyping (NGT) matrix from the H5 file, setting the desired cutoffs for mutation VAF (0.1%) and genotyping call rate (20%). This can take some time to load, so it is recommended to do this once and export the NGT as an RDS file for future use. I have previously exported the NGT matrix as an RDS, so I am just loading it here. H
```{r, NGT extraction, message=FALSE}

# NGT <- read_tapestri_h5_NGT(file,
#        VAF_cutoff = 0.001,
#        GT_cutoff = 20 )

NGT <- readRDS(file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/MRD5_NGT.rds")

```

Filter the NGT matrix based on the recommended cutoffs. This will get rid of low quality genotypes based on read depth (DP), genotype quality score (GQ), and allele frequency (AF). 
```{r,message=FALSE}

#filtered_NGT<-quality_filter_NGT(file=file,
                                # NGT=NGT,
                                # DP_cut=10, 
                                # AF_cut=20,
                                # GQ_cut=30)

filtered_NGT <- readRDS(file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/MRD5_filtered_NGT.rds")

```

Extract the allele frequency (AF) matrix from the .h5 file.
```{r,message=FALSE}

# AF_matrix<- read_tapestri_h5_AF(file)
# AF_matrix_subset<- AF_matrix[rownames(NGT),]


AF_matrix_subset <- readRDS(file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/MRD5_AF_matrix_subset.rds")

```

Extract and filter the AF matrix based on cutoffs described above. 
```{r,message=FALSE}

# filtered_AF <- quality_filter_AF(file=file,
#                                  AF=AF_matrix_subset,
#                                  DP_cut=10, 
#                                  AF_cut=20,
#                                  GQ_cut=30)


filtered_AF <- readRDS(file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/MRD5_filtered_AF.rds")

```

Format data and annotate variants using the `GenomicRanges` and `VariantAnnotation` packages.
```{r,message=FALSE, warning=FALSE}

variants <- rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/id") #all variants
select_variants <- setdiff(colnames(filtered_NGT), "Cell") #variants in the NGT (influenced by VAF and GT cutoffs)

SNV_mat <- data.frame(do.call(cbind, rhdf5::h5read(file = file, name = "/assays/dna_variants/ca/"))) %>%
                      dplyr::filter(.data$id %in% tidyselect::all_of(select_variants)) %>% 
                      dplyr::mutate(ALT = gsub("\\*","N", .data$ALT)) %>% 
                      dplyr::mutate(REF = gsub("\\*","N", .data$REF)) %>% 
                      dplyr::mutate(CHROM = paste0("chr",.data$CHROM)) 

SNV_mat$REF <- as(SNV_mat$REF, "DNAStringSet")
SNV_mat$ALT <- as(SNV_mat$ALT, "DNAStringSet")

variant_gRange <- GenomicRanges::makeGRangesFromDataFrame(SNV_mat, 
                  seqnames.field = "CHROM", start.field = "POS", end.field = "POS", 
                  keep.extra.columns = TRUE)
  
variant_gRange$QUERYID <- 1:length(variant_gRange)

all_variants <- VariantAnnotation::locateVariants(query = variant_gRange, 
                                                          subject = hg19refseq_txdb, region = 
                                                          VariantAnnotation::AllVariants()) %>% 
                                                          data.frame()

coding_variants <- VariantAnnotation::predictCoding(query = variant_gRange, 
                                                            subject = hg19refseq_txdb, 
                                                            seqSource = Hsapiens, varAllele = variant_gRange$ALT) %>%
                                                            data.frame()
                                                            coding_variants$LOCATION<- "coding"                                  
    
intronic_variants <- all_variants %>% 
                     plyranges::filter(as.character(.data$LOCATION) != c("coding")) %>%
                     plyranges::filter(as.character(.data$LOCATION) != c("promoter")) %>%
                     dplyr::inner_join(variant_gRange, by = "QUERYID", copy = TRUE)

spliceSite_variants <- all_variants %>% 
                       plyranges::filter(as.character(.data$LOCATION) == c("spliceSite")) %>%
                       dplyr::inner_join(variant_gRange, by = "QUERYID", copy = TRUE)

splice_intronic <- dplyr::bind_rows(intronic_variants, spliceSite_variants)

out <- dplyr::bind_rows(splice_intronic, coding_variants) %>% 
       dplyr::inner_join(annotation_key) %>% 
       dplyr::mutate(AA = case_when(
       .data$CONSEQUENCE == "nonsynonymous" ~ paste0(.data$gene_name, ".", .data$REFAA, .data$PROTEINLOC,.data$VARAA),
       .data$CONSEQUENCE == "frameshift" ~ paste0(.data$gene_name, ".", .data$REFAA, .data$PROTEINLOC,.data$VARAA),
       .data$CONSEQUENCE == "nonsense" ~ paste0(.data$gene_name, ".", .data$REFAA, .data$PROTEINLOC,.data$VARAA),
       .data$CONSEQUENCE == "synonymous" ~ paste0(.data$gene_name, ".", .data$REFAA, .data$PROTEINLOC,.data$VARAA),
       .data$CONSEQUENCE == "not translated" ~ paste0(.data$gene_name, ".not_txl"),
       .data$LOCATION == "spliceSite" ~ paste0(.data$gene_name, ".splice"),
       .data$LOCATION == "intron" ~ paste0(.data$gene_name, ".intron")))

```

Aggregate the annotated variant information, bulk VAF, genotyping call rate, etc. into a dataframe named `final_mutation_info`.
```{r}

final_mutation_info <- data.frame(Bulk_VAF = apply(filtered_NGT %>% 
                                             dplyr::select(!.data$Cell), MARGIN = 2, function(x) {
                                             (sum(x, na.rm = TRUE)/(sum(!is.na(x)) * 2)) * 100
                                             }), GT_call_rate = apply(filtered_NGT %>% dplyr::select(!.data$Cell), 
                                            MARGIN = 2, function(x) {
                                            100 - (sum(is.na(x))/length(x) * 100)
                                            }), id = colnames(filtered_NGT)[colnames(filtered_NGT) !="Cell"])%>% 
                                            dplyr::inner_join(out, by = "id")

head(final_mutation_info,5)

```

Now we'll pull SNPs from the h.sapiens ENSEMBL SNP mart, and format our list of variants to then identify SNPs that were genotyped in our dataset. This is first achieved below for coding SNPs, which are annotated and then filtered to remove pathogenic variants and SNPs with MAF >0.5. We include the MAF filter to enrich our list for SNPs that are more likely to have divergent genotypes between samples in the multiplex. We end up with a dataframe `coding_SNPs_final`, which has candidate SNPs to use for demultiplexing. 
```{r}

genome <- BSgenome.Hsapiens.UCSC.hg19
          seqlevelsStyle(genome) <- "NCBI"

all_snps <- SNPlocs.Hsapiens.dbSNP144.GRCh37

snp_mart <- useEnsembl(biomart="ENSEMBL_MART_SNP", 
                       host="grch37.ensembl.org", 
                       dataset="hsapiens_snp")

position_coding<- final_mutation_info %>% 
                  dplyr::filter(LOCATION %in% c("coding")) %>%
                  dplyr::select(seqnames,start, end)
                                          
position_coding$seqnames<-gsub("chr","",position_coding$seqnames)
  names(position_coding)[names(position_coding)== "seqnames"]<- "CHR"

position_coding$start<- as.character(position_coding$start)
position_coding$end<- as.character(position_coding$end)

coding_positions <- GPos(seqnames = position_coding$CHR, pos = position_coding$start)
coding_SNPs <- snpsByOverlaps(all_snps, coding_positions, genome = genome)
coding_SNPs_df <- data.frame(chr_pos = start(coding_SNPs), mcols(coding_SNPs)[,c("RefSNP_id", "ref_allele", "alt_alleles")])

coding_SNPs_annotated<-  getBM(attributes=c("refsnp_source",
                                            "refsnp_id",
                                             "chr_name",
                                             "chrom_start",
                                             "chrom_end",
                                             "consequence_type_tv", 
                                             "clinical_significance", 
                                            "minor_allele_freq"	), filters = "snp_filter", 
                                                                   values = coding_SNPs_df$RefSNP_id, 
                                                                   mart = snp_mart)

nonpath_coding_SNPs <- coding_SNPs_annotated[!grepl("pathogenic",coding_SNPs_annotated$clinical_significance),]%>%                       
                       dplyr::filter(is.na(minor_allele_freq ) | minor_allele_freq < 0.5)

coding_SNPs_final <- subset(final_mutation_info, start %in% nonpath_coding_SNPs$chrom_start)

```

This process is repeated to identify nonpathogenic intronic SNPs below. 
```{r}

position_intron <- final_mutation_info %>% dplyr::filter(LOCATION %in% c("intron")) %>%
                                           dplyr::select(seqnames.x,start.x, end.x)

position_intron$seqnames <- gsub("chr","",position_intron$seqnames)
  names(position_intron)[names(position_intron)== "seqnames"]<- "CHR"

position_intron$start <- as.character(position_intron$start)
position_intron$end <- as.character(position_intron$end)


genome <- BSgenome.Hsapiens.UCSC.hg19
all_snps <- SNPlocs.Hsapiens.dbSNP144.GRCh37


seqlevelsStyle(genome) <- "NCBI"

intron_positions <- GPos(seqnames = position_intron$CHR, pos = position_intron$start)

intron_SNPs <- snpsByOverlaps(all_snps, intron_positions, genome = genome)

intron_SNPs_df <- data.frame(chr_pos = start(intron_SNPs), mcols(intron_SNPs)[,c("RefSNP_id", "ref_allele", "alt_alleles")])

intron_SNPs_annotated<-  getBM(attributes=c("refsnp_source",
                                            "refsnp_id",
                                             "chr_name",
                                             "chrom_start",
                                             "chrom_end",
                                             "consequence_type_tv", 
                                             "clinical_significance", 
                                             "minor_allele_freq"	), filters = 'snp_filter', 
                                                                     values = intron_SNPs_df$RefSNP_id, 
                                                                     mart = snp_mart)



intron_SNPs_filtered <- intron_SNPs_annotated[!grepl("pathogenic",intron_SNPs_annotated$clinical_significance),] %>%
                        dplyr::filter(is.na(minor_allele_freq ) | minor_allele_freq < 0.5) 

intron_SNPs_final <- subset(final_mutation_info, start.x %in% intron_SNPs_filtered$chrom_start)

```

Intronic and coding SNPs are then merged, and used to subset the `filtered_AF` dataframe and identify a list of all possible candidate SNPs for demultiplexing. This list will continue to be filtered down, as you will see over the next few chunks of code. 
```{r}

intron_and_coding_SNPs <- rbind(coding_SNPs_final, intron_SNPs_final) %>% 
                          dplyr::filter(!CONSEQUENCE %in% c("nonsense","frameshift")) %>% 
                          dplyr::filter(!duplicated(id))

candidate_SNPs <- filtered_AF %>% 
                  column_to_rownames(., var = "Cell") %>%
                  dplyr::select(c(intron_and_coding_SNPs$id))

```


Since we're concerned with selecting SNPs that are sufficiently divergent between the samples included in the multiplex, we apply an additional filter to only include SNPs which have divergent genotypes (i.e. WT or HET, WT or HOM, HET or HOM ) in a set number of cells, which is defined as:

`n_required` = ((`n_cells_in_multiplexed_run`) x (`n_samples_in_multiplex`)) x (10^-3)

This calculation is based on the reasoning that the fraction of cells that must be divergent in genotype for demultiplexing is proportional to the size of the data set and the number of samples included in the multiplex. Based on an estimated false positive rate of 10^-3 for the TapestriV2 platform, the product of these values is adjusted accordingly. 
```{r}

n_required <- (nrow(filtered_NGT)*5)/1000 #number of cells required to have divergent genotypes

WT_HET_HOM_columns <- apply(candidate_SNPs, 2, function(x){ #divergent genotypes must be WT/HET, or WT/HOM
                            (sum(x < 20, na.rm  = TRUE) >= n_required & (sum(x >= 20  & x <= 80, na.rm  = TRUE) >= n_required |
                            sum(x > 80 & x <= 100, na.rm  = TRUE) >= n_required))}) %>% 
                            as.data.frame() %>% 
                            rename(. = "CONDITION_1") %>% 
                            rownames_to_column(., var = "id")

HET_HOM_columns <- apply(candidate_SNPs, 2, function(x){ #divergent genotypes must beHET/HOM
                         sum(x >= 20  & x <= 80, na.rm  = TRUE) >= n_required & sum(x > 80 & x <= 100, na.rm  = TRUE) >= n_required}) %>% 
                         as.data.frame() %>% 
                         rename(. = "CONDITION_2") %>% 
                         rownames_to_column(., var = "id")
```

After applying these filters, we select the top 20 SNPs with the highest genotyping rate. 
```{r}

#identify SNPs that pass at least 1 condition
satisfied_conditions<- merge(WT_HET_HOM_columns,HET_HOM_columns, by = "id" ) %>% 
                       dplyr::filter(CONDITION_1 == "TRUE" | CONDITION_2 == "TRUE" )  

SNP_list <- final_mutation_info %>% subset(id %in% satisfied_conditions$id) %>%
            arrange(desc(GT_call_rate))

SNPs_to_start<- final_mutation_info %>% subset(id %in% satisfied_conditions$id) %>%
                arrange(desc(GT_call_rate)) %>%
                head(20)

```

We then subset our `filtered_AF` dataframe to select cells which have complete genotyping for these 20 SNPs and store this as a new dataframe, called `complete_20_cells`.
```{r}

AF_matrix_20_SNPs <- column_to_rownames(filtered_AF, var = "Cell") %>% .[SNPs_to_start$id]

tally_perfect_20 <- rowSums(is.na(AF_matrix_20_SNPs)) %>%  #select cells with complete genotyping information for the subset of 20 SNPs
                   as.data.frame() %>%
                   mutate(percent = ./nrow(SNPs_to_start)*100) %>%
                   dplyr::filter(percent <1) 

complete_20_cells <- AF_matrix_20_SNPs[rownames(tally_perfect_20),] #matrix with completely genotyped cells and top 20 SNPs

dim(complete_20_cells) 

```


We then want to exclude extraneous SNPs that won't add value to our analysis. We achieve this by running a principal component analysis and calculating the proportion of variance contributed by each SNP. Applying a filter to select the SNPs that contribute a cumuluative proportion of variance >= 0.99 yields us 14 SNPs to use for demultiplexing. 
```{r}

PCA <- prcomp(complete_20_cells)
prop_var <- PCA$sdev^2 / sum(PCA$sdev^2)
cum_prop_var <- cumsum(prop_var)
n_components <- which(cum_prop_var >= .99)[1]
top_PCS <- PCA$rotation[1:n_components,] %>% as.data.frame()

nrow(top_PCS) #14 SNPs
```


We then re-subset our starting allele frequency data (`filtered_AF`) on cells with complete genotyping for these 14 SNPs. This is stored as `clustering_AF_matrix`.
```{r}

AF_matrix_SNPs <- column_to_rownames(filtered_AF, var = "Cell") %>% .[rownames(top_PCS)]

Cell_NA_tally <- rowSums(is.na(AF_matrix_SNPs)) %>%  #select cells with perfect genotyping information for the subset of SNPs
                 as.data.frame() %>%
                 mutate(percent = ./nrow(top_PCS)*100) %>%
                 dplyr::filter(percent <1) 

clustering_AF_matrix <- AF_matrix_SNPs[rownames(Cell_NA_tally),]
dim(clustering_AF_matrix) #1,398 cells with complete genotyping for the 14 SNPs

```

Next, we run K-means clustering with the `clustering_AF_matrix`, setting the number of centers equal to the number of unique samples within the multiplex (in this case, 5 samples). For visualization purposes, we run UMAP analysis on SNP allele frequencies, with n_neighbors set to the square root of the total number of cells in the `clustering_AF_matrix`. 
```{r}

set.seed(1)

kmeans_AF <- kmeans(clustering_AF_matrix, centers = 5, nstart = 30) #centers set based on number of samples in the multiplex 

umap_SNPs <-umap(clustering_AF_matrix, n_neighbors = sqrt(nrow(clustering_AF_matrix))) 

cell_cluster <- kmeans_AF$cluster %>% as.data.frame() %>% rename(. = "Cluster")

umap_df <- umap_SNPs$layout %>%
           as.data.frame() %>% 
           merge(., cell_cluster, by = 0) %>% 
           column_to_rownames(., var = "Row.names")

names(umap_df)[names(umap_df)== "V1"] <- "UMAP1"
names(umap_df)[names(umap_df)== "V2"] <- "UMAP2"
umap_df$Cluster <- as.character(umap_df$Cluster)

 umap_df %>% ggplot(aes(x = UMAP1, y = UMAP2))+
             geom_point(aes( fill = Cluster), color = "black",alpha = .2, size =2, shape =21, show.legend = TRUE)+
             scale_fill_manual(values = rev(c(tol(5))))+
                               theme_classic()+
                               theme(axis.title.x = element_text(size = 12),
                                     axis.title.y = element_text(size = 12),
                                     axis.text.x = element_text(size = 10),
                                     axis.text.y = element_text(size = 10),
                                     legend.text = element_text(size =10),
                                     plot.title = element_text(hjust = 0.5, size = 14))+
                                labs(x = "UMAP1",y = "UMAP2")


```

Now we'll take a look at how well the SNPs are separating into each cluster. We can generate a SNP-cluster heatmap using the Superheat package.
```{r}

cluster_df <- kmeans_AF$cluster %>% as.data.frame() %>% rename("." = "Cluster")

AF_cluster <- filtered_AF %>% column_to_rownames(.,  var ="Cell") %>% 
              merge(., cluster_df, by=0, all=FALSE) %>% 
              column_to_rownames(., var = "Row.names") %>% 
              .[c(colnames(clustering_AF_matrix),"Cluster")]

#x-axis has SNPs, y axis has cluster
superheat(AF_cluster[,1:ncol(AF_cluster)-1], 
          scale = FALSE,
          bottom.label.text.angle = 90,
          membership.rows = AF_cluster$Cluster,
          bottom.label.size = 0,
          bottom.label.text.size = 2.5,
          left.label.text.size = 2.8,
          left.label.size = .1,
          pretty.order.rows = TRUE,
          pretty.order.cols = TRUE, 
          smooth.heat = FALSE)

```

While the SNPs appear to segregate well into distinct clusters, we know that within this dataset there exist 1) doublets, 2) genotypes that are driven by allele dropout (ADO) and/or imperfect sequencing. We'll now start the process of filtering out these difficult-to-classify cells. First, we calculate the number of cells currently assigned to each cluster.
```{r}

cluster_ncell <- AF_cluster %>% group_by(Cluster) %>% summarise(n_cells = n()) 

cluster_ncell

```

Next, we evenly sample each cluster and create a pool of cells with even representation from each cluster. The number of cells to sample is set by the cluster with the lowest number of cells. 
```{r}
set.seed(1)

AF_cluster_list <- AF_cluster %>% split(AF_cluster$Cluster)

sampled_pool_list <- lapply(AF_cluster_list,  function(df, n = min(cluster_ncell$n_cells)){ 
                                                          sampled_cells <- df %>% 
                                                                dplyr::select(!c("Cluster")) %>%
                                                                .[sample(nrow(.), n), ] 
                                              return(sampled_cells)
                                          }) 


sampled_pool <- sampled_pool_list %>% do.call(rbind,.) 

dim(sampled_pool)
```

From this pool of cells, we will then generate artificial doublets via randomly sampling 2 cells at a time and averaging their SNP allele frequencies. Below are two functions which let us sample and average SNP profiles. 
```{r}

#functions to sample and mix cells
mix_cluster_AF <- function(df){
                  df <- df
                  df$mixed <- (df[,1]+ df[,2])/2
                  names(df)[names(df) == "mixed"] <- paste0(colnames(df[1:2]), collapse="_")
                      return(df)
      }

mix_cells <- function(AF_df, n){
             sample <- sample_n(AF_df,2, replace = FALSE) %>% t() %>% as.data.frame()
             artificial_mix <- combn(sample, 2, simplify = FALSE)
             cell_mixes <- lapply(artificial_mix, mix_cluster_AF) %>% 
                          bind_cols()%>% 
                          dplyr::select(.,contains("_")) %>% t() %>% as.data.frame()
             return(cell_mixes)
}

```
 
We'll generate enough doublets to make up roughly 5-10% of the total dataset. We start by simulating doublets at 10% of the total data. After removing homotypic doublets (i.e. doublets created by two cells from the same cluster), we end up with 486 artificial heterotypic doublets
((486/5938) *100 = 8.18%).
```{r}

set.seed(1)

sampled_all <- replicate(n = nrow(filtered_NGT)*.1, mix_cells(sampled_pool), simplify = FALSE) %>% bind_rows() #10% of total data

sampled_all_heterotypic <- sampled_all %>% 
                            mutate(type = gsub("[^0-9]+", "", rownames(.))) %>% subset(!type == "11" & !type == "22" &
                                                                                       !type == "33" & !type == "44" &
                                                                                       !type == "55") %>% dplyr::select(!c(type))

nrow(sampled_all_heterotypic)/nrow(filtered_NGT)*100

```


We then generate a cluster center for the artificial doublets via K-means clustering with a center value of 1. The artificial doublets are then combined with the real clusters generated previously, and we label them as artificial or real cells, respectively. 
```{r}
set.seed(1)

kmeans_sampled <- kmeans(sampled_all_heterotypic, centers = 1, nstart = 30)
kmeans_sampled_center <- kmeans_sampled$centers[1,]

kmeans_sampled_clusters <- kmeans_sampled$cluster %>% as.data.frame() %>% rename(. = "Cluster")
  kmeans_sampled_clusters$Cluster[kmeans_sampled_clusters$Cluster == "1"] <- "artificial"

real_and_mixed_clusters <- rbind(kmeans_sampled_clusters, cell_cluster) #bind mix cluster with real clusters
real_and_mixed_clusters$Cluster[real_and_mixed_clusters$Cluster == "1" | 
                                real_and_mixed_clusters$Cluster == "2" |
                                real_and_mixed_clusters$Cluster == "3" |
                                real_and_mixed_clusters$Cluster == "4" |
                                real_and_mixed_clusters$Cluster == "5" ] <- "real"
```

We can then generate a UMAP projection of real and artificial cells together. 
```{r}

cells_and_sampled_AF <- rbind(clustering_AF_matrix, sampled_all_heterotypic) 

umap_SNPs_real_and_mixed <-umap(cells_and_sampled_AF, n_neighbors = 5)

umap_df_real_and_mixed <- umap_SNPs_real_and_mixed$layout %>%
                          as.data.frame() %>% 
                          merge(., real_and_mixed_clusters, by = 0) %>% 
                          column_to_rownames(., var = "Row.names")

names(umap_df_real_and_mixed)[names(umap_df_real_and_mixed)== "V1"] <- "UMAP1"
names(umap_df_real_and_mixed)[names(umap_df_real_and_mixed)== "V2"] <- "UMAP2"
umap_df_real_and_mixed$Cluster <- as.character(umap_df_real_and_mixed$Cluster)

umap_real_mixed<- umap_df_real_and_mixed %>% ggplot(aes(x = UMAP1, y = UMAP2))+
                  geom_point(aes( fill = Cluster), color = "black",alpha = .2, size =2, shape =21, show.legend = TRUE)+
                            scale_color_manual(values = c("artificial"="black","real"="black"))+  
                            scale_fill_manual(values = c("artificial"="#EA3546","real"="#00A7E1"))+                
                              theme_classic()+
                              theme(axis.title.x = element_text(size = 12),
                                        axis.title.y = element_text(size = 12),
                                        axis.text.x = element_text(size = 10),
                                        axis.text.y = element_text(size = 10),
                                        legend.text = element_text(size = 10),
                                        plot.title = element_text(hjust = 0.5, size = 14))+
                            labs(x = "UMAP1",y = "UMAP2")

umap_real_mixed


```

Next we will calculate the Euclidean distance between reals cells and the artificial center using the pdist function within the `pdist` package. 
```{r, message=FALSE, warning=FALSE}

dist_C1_mix <- pdist(AF_cluster_list[1]$`1`[1:ncol(AF_cluster)-1], kmeans_sampled_center) %>% 
               as.matrix() %>% as.data.frame() %>% rename("V1"= "mix")

dist_C2_mix <- pdist(AF_cluster_list[2]$`2`[1:ncol(AF_cluster)-1], kmeans_sampled_center) %>% 
               as.matrix() %>% as.data.frame() %>% rename("V1"= "mix")

dist_C3_mix <- pdist(AF_cluster_list[3]$`3`[1:ncol(AF_cluster)-1], kmeans_sampled_center) %>% 
               as.matrix() %>% as.data.frame() %>% rename("V1"= "mix")

dist_C4_mix <- pdist(AF_cluster_list[4]$`4`[1:ncol(AF_cluster)-1], kmeans_sampled_center) %>% 
               as.matrix() %>% as.data.frame() %>% rename("V1"= "mix")

dist_C5_mix <- pdist(AF_cluster_list[5]$`5`[1:ncol(AF_cluster)-1], kmeans_sampled_center) %>% 
               as.matrix() %>% as.data.frame() %>% rename("V1"= "mix")

rownames(dist_C1_mix) <- rownames(AF_cluster_list[1]$`1`)
rownames(dist_C2_mix) <- rownames(AF_cluster_list[2]$`2`)
rownames(dist_C3_mix) <- rownames(AF_cluster_list[3]$`3`)
rownames(dist_C4_mix) <- rownames(AF_cluster_list[4]$`4`)
rownames(dist_C5_mix) <- rownames(AF_cluster_list[5]$`5`)

distances_cluster_mix <- rbind(dist_C1_mix, dist_C2_mix) %>%
                         rbind(., dist_C3_mix) %>%
                         rbind(., dist_C4_mix) %>%
                         rbind(., dist_C5_mix)

```


We'll also calculate the Euclidean distance from real cells to their respective real centers, which we pull from the K-means clustering output. Here we are asking: if a cell is partitioned into a cluster, how similar is that cell's SNP profile to the average SNP profile of that cluster?
```{r}

C1_mean_AF <- kmeans_AF$centers[1,]
C2_mean_AF <- kmeans_AF$centers[2,]
C3_mean_AF <- kmeans_AF$centers[3,]
C4_mean_AF <- kmeans_AF$centers[4,]
C5_mean_AF <- kmeans_AF$centers[5,]

dist_C1_real <- pdist(AF_cluster_list[1]$`1`[1:ncol(AF_cluster)-1], C1_mean_AF) %>% 
                as.matrix() %>% as.data.frame() %>% rename("V1"= "real")

dist_C2_real <- pdist(AF_cluster_list[2]$`2`[1:ncol(AF_cluster)-1], C2_mean_AF) %>% 
                as.matrix() %>% as.data.frame() %>% rename("V1"= "real")

dist_C3_real <- pdist(AF_cluster_list[3]$`3`[1:ncol(AF_cluster)-1], C3_mean_AF) %>% 
                as.matrix() %>% as.data.frame() %>% rename("V1"= "real")

dist_C4_real <- pdist(AF_cluster_list[4]$`4`[1:ncol(AF_cluster)-1], C4_mean_AF) %>% 
                as.matrix() %>% as.data.frame() %>% rename("V1"= "real")

dist_C5_real <- pdist(AF_cluster_list[5]$`5`[1:ncol(AF_cluster)-1], C5_mean_AF) %>% 
                as.matrix() %>% as.data.frame() %>% rename("V1"= "real")

rownames(dist_C1_real)<- rownames(AF_cluster_list[1]$`1`)
rownames(dist_C2_real)<- rownames(AF_cluster_list[2]$`2`)
rownames(dist_C3_real)<- rownames(AF_cluster_list[3]$`3`)
rownames(dist_C4_real)<- rownames(AF_cluster_list[4]$`4`)
rownames(dist_C5_real)<- rownames(AF_cluster_list[5]$`5`)

distances_cluster_real <- rbind(dist_C1_real, dist_C2_real) %>%
                          rbind(., dist_C3_real) %>%
                          rbind(., dist_C4_real) %>%
                          rbind(., dist_C5_real)

```

We then want to calculate how close 95% of completely-genotyped cells are to their respective cluster center. This will be used as a cutoff to filter out cells that fall within this distance to the artificial (mixed) center.
```{r}

quantile_95 <- quantile(distances_cluster_real$real, probs = c(0.95)) 

quantile_95

```

We can plot these measurements below.
```{r}
real_dist_histogram<- ggplot(distances_cluster_real, aes(x=real)) +
                      geom_density(color = "black", fill = "black")+
                      theme_bw()+
                       annotate(geom = "rect", xmin = 0, xmax = quantile_95, ymin = 0, ymax = 0.035,
                               fill = "#00A7E1", colour = "black", alpha = 0.4, linetype = "dashed") +
                      xlim(0,200)+
                      ylim(0, 0.035)+
                      xlab("Euclidean Distance")+
                      ylab("Density")+
                      theme(axis.title.x = element_text(size = 12),
                                        axis.title.y = element_text(size = 12),
                                        axis.text.x = element_text(size = 10),
                                        axis.text.y = element_text(size = 10),
                                        legend.text = element_text(size = 10),
                                        plot.title = element_text(hjust = 0.5, size = 14))+
                      ggtitle("Distribution of Complete Cells Distance to Real Center")+
                      theme(plot.title = element_text(hjust = 0.5, size = 14), legend.title = element_blank())

real_dist_histogram

```

```{r}

distances_cluster_real_df<- merge(distances_cluster_real, cell_cluster, by = 0) %>% 
                            column_to_rownames(., var = "Row.names")


vln_real_dist <- ggplot(distances_cluster_real_df, aes(x = real, y = factor(Cluster), fill = factor(Cluster))) +
                 scale_fill_manual(values = rev(c(tol(5))))+
                 theme_bw()+
                 geom_violin(alpha = 0.5) +
                 geom_jitter( height = 0.2, size = .01)+
                    theme(axis.title.x = element_text(size = 12),
                          axis.title.y = element_text(size = 12),
                          axis.text.x = element_text(size = 10),
                          axis.text.y = element_text(size = 10),
                          legend.text = element_text(size = 10),
                          plot.title = element_text(hjust = 0.5, size = 14))+
                ggtitle("Euclidean Distance of Real Cells to Real Cluster Centers")+
                xlab("Euclidean Distance")+
                ylab("Cluster")+
                theme(legend.position = "none")

vln_real_dist

```

```{r}

distances_cluster_mix_df<- merge(distances_cluster_mix, cell_cluster, by = 0) %>% 
                              column_to_rownames(., var = "Row.names")

labels_hist<- c("Cluster 1", "Cluster 2", "Cluster 3","Cluster 4", "Cluster 5")
names(labels_hist) <- c("1", "2", "3","4","5")

dist_mix_hist<- ggplot(distances_cluster_mix_df, aes(x=mix, fill=factor(Cluster))) +
                geom_density(alpha=.5, show.legend = FALSE)+
                theme_bw()+
                annotate(geom = "rect", xmin = 0, xmax = quantile_95, ymin = 0, ymax = 0.06,
                         fill = "black", colour = "black", alpha = 0.3, linetype = "dashed") +
                         scale_fill_manual(values = rev(c(tol(5))))+
                                  xlim(0,200)+
                                  xlab("Euclidean Distance")+
                                  ylab("Density")+
                                  theme(axis.title.x = element_text(size = 12),
                                                      axis.title.y = element_text(size = 12),
                                                      axis.text.x = element_text(size = 9),
                                                      axis.text.y = element_text(size = 10),
                                                      legend.text = element_text(size = 10),
                                                      plot.title = element_text(hjust = 0.5, size = 14))+
                                  theme(plot.title = element_text(hjust = 0.5, size = 14), 
                                        legend.title = element_blank(), strip.background = element_rect(
                                        color="black", fill="white", size=0.5, linetype="solid"))+
                                  facet_wrap(~Cluster, labeller = labeller(Cluster = labels_hist))

dist_mix_hist
#ggsave(dist_mix_hist,width = 5, height = 3, dpi = 500, filename = "real_cell_art_cluster.tiff", device = "tiff")

```

After checking our data, we then want to start the doublet removal process. We'll now select all cells and the the subset of SNPs we previously used for clustering.
```{r}

#this data frame containing imperfect genotypes will be used as input to filter out putative doublets from the data set in the next few chunks of code.
imperfect_AF <- filtered_AF %>% 
              .[,c("Cell",colnames(clustering_AF_matrix))]  

table(is.na(imperfect_AF))

```

Since we ultimately want to get an estimate of the distance from all cells to the real and artificial cluster centers, we need to get rid of NAs for SNPs with missing genotyping information. Below are functions to randomly sample and replace NAs with AF values of 0 (wild type), 50 (heterozygous), or 100 (homozygous). 
```{r}

lets_sample <- function(AF){
                  SNP_AF <- AF
                  sampled_AF <- as.data.frame(apply(SNP_AF,2, function(x){
                                              ifelse((is.na(x)),
                                              sample(c(0,50,100), 
                                              replace = TRUE),x)})) 
                  return(sampled_AF)
                      }


sample_for_mixing <- function(df, ref, center, quantile){
                        df_sampled <- lets_sample(AF = df)
                        df_sampled <- df_sampled[,colnames(ref)]
                        
                        distance <- pdist(df_sampled, center) %>% 
                                          as.matrix() %>% 
                                          as.data.frame() %>% 
                                          rename("V1"= "mix_dist") 
                        
                        rownames(distance) <- rownames(df_sampled)
                        
                        distance_filtered <- distance %>% 
                                               dplyr::filter(mix_dist > quantile)
                        
                        return(distance_filtered)
                        
                        }
```

Replicate the process 10 times, and then select the cells that survive the doublet filter through all replicates. The survivors represent the pool of cells that we will classify later on. 
```{r, message=FALSE, warning=FALSE}

set.seed(1)
replicated_sampling <- replicate(10, sample_for_mixing(df = (imperfect_AF %>%
                                                            column_to_rownames(., var = "Cell")),
                                 ref = clustering_AF_matrix,
                                 center = kmeans_sampled_center,
                                 quantile = quantile_95) %>% 
                                 rownames_to_column(., var = "Cell"))

survivors <- merge(replicated_sampling[,1], replicated_sampling[,2], by = "Cell") %>%
             merge(., replicated_sampling[,3], by = "Cell") %>%
             merge(., replicated_sampling[,4], by = "Cell") %>%
             merge(., replicated_sampling[,5], by = "Cell") %>%
             merge(., replicated_sampling[,6], by = "Cell") %>%
             merge(., replicated_sampling[,7], by = "Cell") %>%
             merge(., replicated_sampling[,8], by = "Cell") %>%
             merge(., replicated_sampling[,9], by = "Cell") %>%
             merge(., replicated_sampling[,10], by = "Cell") %>% 
             column_to_rownames(., var = "Cell")

```

```{r}
dim(survivors)
```


In the above code, we filtered out doublets in the total dataset to obtain a pool of cells that to classify later. Before we do that, we need to identify the SNP profile of each cluster, which we'll then use to classify the remaining cells. Current clusters have doublets present, so we still need to filter out doublets in this subset of data. 
```{r}

clusters_filtered<- cbind(distances_cluster_real, distances_cluster_mix) %>% 
                    dplyr::filter(mix > quantile_95) %>% #dist to artificial center must be larger than quantile_95
                    dplyr::filter(mix > real ) #cell must be closer to its own center than to the artificial center

clusters_singlets<- clustering_AF_matrix[rownames(clusters_filtered), colnames(clustering_AF_matrix)]

dim(clusters_singlets)
```

Now we'll recluster the subset of completely genotyped cells with doublets removed. As we did earlier on, we set k=5 since there are 5 unique samples in this multiplexed run. 
```{r}

set.seed(1)

km_singlets<- kmeans(clusters_singlets, centers = 5, nstart = 30)

umap_singlets<-umap(clusters_singlets, n_neighbors = sqrt(nrow(clusters_singlets))) 
cell_cluster_singlets<- km_singlets$cluster %>% as.data.frame() %>% rename(. = "Cluster")

umap_df_singlets<- umap_singlets$layout %>%
                   as.data.frame() %>% 
                   merge(., cell_cluster_singlets, by = 0) %>% 
                   column_to_rownames(., var = "Row.names")


names(umap_df_singlets)[names(umap_df_singlets)== "V1"]<-"UMAP1"
names(umap_df_singlets)[names(umap_df_singlets)== "V2"]<-"UMAP2"
umap_df_singlets$Cluster<- as.character(umap_df_singlets$Cluster)

umap_singlets_gg<-  umap_df_singlets %>% ggplot(aes(x = UMAP1, y = UMAP2))+
                      geom_point(aes( fill = Cluster), color = "black",alpha = .2, size =2, shape =21, show.legend = TRUE)+
                      scale_fill_manual(values = rev(c(tol(5))))+
                  theme_classic()+
                theme(plot.title = element_text(hjust = 0.5, size = 14))+
                labs(x = "UMAP1",
                     y = "UMAP2")

umap_singlets_gg

#ggsave(umap_singlets_gg, width = 3, height = 2.2, dpi = 500, filename = "MRD5_UMAP3.tiff", device = "tiff")

```
Again we'll inspect the SNP profile of each cluster
```{r}

singlet_cluster_df<- km_singlets$cluster %>% as.data.frame() %>% rename(. = "Cluster")

NGT_cluster_singlets<- filtered_NGT %>%
                      column_to_rownames(.,  var ="Cell") %>% 
                      merge(.,singlet_cluster_df, by = 0, all =FALSE ) %>%
                      column_to_rownames(., var = "Row.names") %>% 
                    .[c(colnames(clusters_singlets),"Cluster")]

#x-axis has SNPs, y-axis has cluster
superheat(NGT_cluster_singlets[, 1:ncol(NGT_cluster_singlets)-1],
          scale = FALSE,
          bottom.label.text.angle = 90,
          membership.rows = NGT_cluster_singlets$Cluster,
          bottom.label.size = 0,
          left.label.text.size = 2.8,
          left.label.size = .1,
          pretty.order.rows = TRUE,
          pretty.order.cols = TRUE, 
          smooth.heat = FALSE)

ordered_singlets<- NGT_cluster_singlets %>% arrange(Cluster)

```

```{r}

#this code generates a heatmap in Fig 2 in the manuscript
    # pheatmap(ordered_singlets[, 1:ncol(ordered_singlets)-1], 
    #          show_rownames = FALSE, 
    #          show_colnames = FALSE,
    #          cluster_rows = FALSE, 
    #          cluster_cols = FALSE, 
    #          border_color = NA,
    #          legend = FALSE,
    #          color = c("#E5DADA","#840032","#002642"))

```
Now we'll prepare the data for classification i.e. matching of cells to clusters based on SNP profiles. 
```{r}

#pull surviving cells after de-doubleting and subset on the 14 SNPs. 
NGT_to_classify<- filtered_NGT %>% 
                  column_to_rownames(., var = "Cell") %>% 
                  .[rownames(survivors), colnames(clustering_AF_matrix)] %>%
                  apply(., 1, as.character)

rownames(NGT_to_classify)<- colnames(clustering_AF_matrix)

matching_df<- NGT_to_classify %>% as.data.frame()

#convert data to character
matching_NGT<- apply(matching_df, 2, as.character) %>% as.data.frame()
              
rownames(matching_NGT)<- rownames(matching_df)

#split into multiple dataframes
matching_NGT_split<- split.default(matching_NGT, c(1:ncol(matching_NGT))) 

#get rid of columns with NAs, as we don't want to penalize cells with missing SNP info during matching
matching_df_split_noNA<- lapply(matching_NGT_split, na.omit)

```

This function calculates the Hamming distance between each cells SNP profile and the cluster reference profile that is supplied. 
```{r}

match_cells <- function(cluster, x, names ){
                        merged<-lapply(x, function(y) merge(y,cluster, by = 0 ))
                        matching_index<- lapply(merged, function(x){sum(x[,2] != x[,3])})
                        matched_df<- as.data.frame(do.call(rbind,matching_index))
                        rownames(matched_df)<- colnames(names)
               return(matched_df)
              }

```

Here we prepare the de-doubleted data to determine the most common genotype for each SNP in each cluster. This is done with NGT values (0, 1, 2). 
```{r}

singlets_AF<- clustering_AF_matrix[rownames(singlet_cluster_df),] %>% 
              cbind(., singlet_cluster_df) %>% 
              split(.$Cluster) 

C1_singlets<- singlets_AF$`1`[1:ncol(clustering_AF_matrix)]
C2_singlets<- singlets_AF$`2`[1:ncol(clustering_AF_matrix)]
C3_singlets<- singlets_AF$`3`[1:ncol(clustering_AF_matrix)]
C4_singlets<- singlets_AF$`4`[1:ncol(clustering_AF_matrix)]
C5_singlets<- singlets_AF$`5`[1:ncol(clustering_AF_matrix)]

C1_to_match<- matching_df[,rownames(C1_singlets)] %>% apply(., 2, as.character) %>% as.data.frame()
C2_to_match<- matching_df[,rownames(C2_singlets)] %>% apply(., 2, as.character) %>% as.data.frame()
C3_to_match<- matching_df[,rownames(C3_singlets)] %>% apply(., 2, as.character) %>% as.data.frame()
C4_to_match<- matching_df[,rownames(C4_singlets)] %>% apply(., 2, as.character) %>% as.data.frame()
C5_to_match<- matching_df[,rownames(C5_singlets)] %>% apply(., 2, as.character) %>% as.data.frame()

  rownames(C1_to_match)<- colnames(C1_singlets)
  rownames(C2_to_match)<- colnames(C2_singlets)
  rownames(C3_to_match)<- colnames(C3_singlets)
  rownames(C4_to_match)<- colnames(C4_singlets)
  rownames(C5_to_match)<- colnames(C5_singlets)

```


Determine the most common genotype for each SNP in each cluster, again using NGT values. 
```{r, match cells to clusters}

C1_input<- apply(C1_to_match,1,function(x) names(which.max(table(x)))) %>% as.data.frame() %>% rename("." = "C1")
C2_input<- apply(C2_to_match,1,function(x) names(which.max(table(x)))) %>% as.data.frame() %>% rename("." = "C2")
C3_input<- apply(C3_to_match,1,function(x) names(which.max(table(x)))) %>% as.data.frame() %>% rename("." = "C3")
C4_input<- apply(C4_to_match,1,function(x) names(which.max(table(x)))) %>% as.data.frame() %>% rename("." = "C4")
C5_input<- apply(C5_to_match,1,function(x) names(which.max(table(x)))) %>% as.data.frame() %>% rename("." = "C5")

input_df<- cbind(C1_input, C2_input) %>%
           cbind(., C3_input) %>%
           cbind(., C4_input) %>%
           cbind(., C5_input) %>% t() %>%
           as.data.frame() %>% 
           apply(., 2, as.numeric) %>% 
           as.data.frame()


```


```{r}
pheatmap(input_df[, 1:ncol(input_df)], 
         show_rownames = TRUE, 
         show_colnames = FALSE,
         cluster_rows = FALSE, 
         cluster_cols = TRUE, 
         border_color = "black",
         legend = TRUE, 
         gaps_row = c(1,2,3,4),
         color = c("#E5DADA","#840032","#002642"))


```

Match cells to each cluster reference using the `match_cells` function, which requires a cluster reference profile, a list containing SNP profiles for all cells to be classified, and a data frame to supply barcode identities. After running this function over each cluster, each barcode will have a Hamming distance to each cluster.  
```{r}

C1_matched<- match_cells(cluster = C1_input, x = matching_df_split_noNA, names = matching_NGT)
C2_matched<- match_cells(cluster = C2_input, x = matching_df_split_noNA, names = matching_NGT)
C3_matched<- match_cells(cluster = C3_input, x = matching_df_split_noNA, names = matching_NGT)
C4_matched<- match_cells(cluster = C4_input, x = matching_df_split_noNA, names = matching_NGT)
C5_matched<- match_cells(cluster = C5_input, x = matching_df_split_noNA, names = matching_NGT)

```

Now we'll calculate the Hamming distance between the clusters themselves. This will ensure that if there are clusters that have highly similar SNP profiles (i.e. Hamming distance <3) we can account for it in the downstream filtering steps. 
```{r}

cluster_input <- cbind(C1_input, C2_input) %>%
                 cbind(., C3_input) %>%
                 cbind(., C4_input) %>%
                 cbind(., C5_input) %>% t() %>% as.data.frame()

cluster_input_hamming <- as.data.frame(as.matrix(hammingdists(cluster_input)))

rownames(cluster_input_hamming)<- rownames(cluster_input)
colnames(cluster_input_hamming)<- rownames(cluster_input)

superheat(cluster_input_hamming,
              scale = FALSE,
              left.label.text.size = 3,
              pretty.order.rows = TRUE,
              pretty.order.cols = TRUE, 
              left.label.size = .12,
              bottom.label.size = .09,
              bottom.label.text.size = 3,
              legend = FALSE,
              X.text = round(as.matrix(cluster_input_hamming), 1))

```
The Hamming distance data are then merged into a single dataframe. We also want to calculate the amount of missing SNPs in each cell that was included in cell matching; this will enable us to calculate the number of possible SNP matches between cell and cluster vs. the number of matches we observed. 
```{r}

available_info_singlets <- filtered_NGT %>% 
                           column_to_rownames(., var = "Cell") %>% 
                          .[rownames(survivors), colnames(input_df)] %>%
                           mutate(prop_avail_info = (((ncol(.) - rowSums(is.na(.)))/ncol(.)))) %>%
                           dplyr::select(c(prop_avail_info))

C1_info <- C1_matched %>% rename("V1"= "C1")
C2_info <- C2_matched %>% rename("V1"= "C2")
C3_info <- C3_matched %>% rename("V1"= "C3")
C4_info <- C4_matched %>% rename("V1"= "C4")
C5_info <- C5_matched %>% rename("V1"= "C5")

match_master <- cbind(C1_info, C2_info) %>%
                cbind(., C3_info) %>%
                cbind(., C4_info) %>%
                cbind(., C5_info) %>% 
                cbind(.,available_info_singlets) %>%
                mutate(possible_matches = (prop_avail_info * ncol(input_df))) %>%
                mutate(num_missing_SNPs = (ncol(input_df) - possible_matches)) %>%
                mutate(C1_abs = (possible_matches - C1)) %>%
                mutate(C2_abs = (possible_matches - C2)) %>%
                mutate(C3_abs = (possible_matches - C3)) %>%
                mutate(C4_abs = (possible_matches - C4)) %>%
                mutate(C5_abs = (possible_matches - C5))

head(match_master)
```

Cells are then assigned to clusters based on the available SNP information matching perfectly to one cluster, while being a Hamming distance of 3 to every other cluster. For some multiplexed runs, the maximum Hamming distance between clusters was 2, in which case we reduced the classification filter to accommodate this. In scMRD run MRD5, none of the clusters had a Hamming distance <3. 
```{r}

C1_filtered<- match_master %>% 
              dplyr::filter(C1_abs >= possible_matches)  %>% #must match perfectly to cluster 1
              dplyr::filter((C1_abs - C2_abs) >= 3 &  #must be Hamming distance of >=3 to cluster 2-5
                            (C1_abs - C3_abs) >= 3 & 
                            (C1_abs - C4_abs) >= 3 &
                            (C1_abs - C5_abs) >= 3)

C2_filtered<- match_master %>% 
              dplyr::filter(C2_abs >= possible_matches)  %>%
              dplyr::filter((C2_abs - C1_abs) >= 3 & 
                            (C2_abs - C3_abs) >= 3 & 
                            (C2_abs - C4_abs) >= 3 &
                            (C2_abs - C5_abs) >= 3)

C3_filtered<- match_master %>% 
              dplyr::filter(C3_abs >= possible_matches)  %>%
              dplyr::filter((C3_abs - C1_abs) >= 3 & 
                            (C3_abs - C2_abs) >= 3 & 
                            (C3_abs - C4_abs) >= 3 &
                            (C3_abs - C5_abs) >= 3)  

C4_filtered<- match_master %>% 
              dplyr::filter(C4_abs >= possible_matches)  %>%
              dplyr::filter((C4_abs - C1_abs) >= 3 & 
                            (C4_abs - C2_abs) >= 3 & 
                            (C4_abs - C3_abs) >= 3 &
                            (C4_abs - C5_abs) >= 3)

C5_filtered<- match_master %>% 
              dplyr::filter(C5_abs >= possible_matches)  %>%
              dplyr::filter((C5_abs - C1_abs) >= 3 & 
                            (C5_abs - C2_abs) >= 3 & 
                            (C5_abs - C3_abs) >= 3 &
                            (C5_abs - C4_abs) >= 3)
                              
```


Subset the starting NGT data (`filtered_NGT`) for cells classified to each cluster. 
```{r}

C1_final<-subset(filtered_NGT, Cell %in% rownames(C1_filtered),)
C2_final<-subset(filtered_NGT, Cell %in% rownames(C2_filtered),)
C3_final<-subset(filtered_NGT, Cell %in% rownames(C3_filtered),)
C4_final<-subset(filtered_NGT, Cell %in% rownames(C4_filtered),)
C5_final<-subset(filtered_NGT, Cell %in% rownames(C5_filtered),)

```

```{r}
nrow(C1_final)
nrow(C2_final)
nrow(C3_final)
nrow(C4_final)
nrow(C5_final)
```

Now that we've binned cells into clusters, we'll detect mutations present in each of the clusters. Below is a function to annotate variants, and calculate VAF + genotyping call rate using a "Cx_final" NGT matrix as input. 
```{r}

detect_mutations <- function(NGT) { 
                 mutations <- data.frame(Bulk_VAF = apply(NGT %>% 
                              dplyr::select(!.data$Cell), MARGIN = 2, function(x) {
                              (sum(x, na.rm = TRUE)/(sum(!is.na(x)) * 2)) * 100
                              }), GT_call_rate = apply(NGT %>% dplyr::select(!.data$Cell), 
                              MARGIN = 2, function(x) {100 - (sum(is.na(x))/length(x) * 100)
                              }), id = colnames(NGT)[colnames(NGT) != "Cell"]) %>% 
                              dplyr::inner_join(out, by = "id") %>%
                             dplyr::select(Bulk_VAF, GT_call_rate, AA, id, CONSEQUENCE)
              return(mutations)
  }

```

```{r}

Cluster_check_1_matched <- detect_mutations(C1_final)

Cluster_check_filter_1_matched<- Cluster_check_1_matched %>%
                                 dplyr::filter(GT_call_rate > 10) %>%
                                 dplyr::select(Bulk_VAF, id)

names(Cluster_check_filter_1_matched)[names(Cluster_check_filter_1_matched) == "Bulk_VAF"]<-"C1_VAF"

```

```{r}
Cluster_check_2_matched <- detect_mutations(C2_final)

Cluster_check_filter_2_matched<- Cluster_check_2_matched %>%
                                 dplyr::filter(GT_call_rate > 10) %>%
                                 dplyr::select(Bulk_VAF, id)

names(Cluster_check_filter_2_matched)[names(Cluster_check_filter_2_matched) == "Bulk_VAF"]<-"C2_VAF"

```

```{r}
Cluster_check_3_matched <- detect_mutations(C3_final)

Cluster_check_filter_3_matched<- Cluster_check_3_matched %>%
                                 dplyr::filter(GT_call_rate > 10) %>%
                                 dplyr::select(Bulk_VAF, id)

names(Cluster_check_filter_3_matched)[names(Cluster_check_filter_3_matched) == "Bulk_VAF"]<-"C3_VAF"

```

```{r}
Cluster_check_4_matched <- detect_mutations(C4_final)

Cluster_check_filter_4_matched<- Cluster_check_4_matched %>%
                                 dplyr::filter(GT_call_rate > 10) %>%
                                 dplyr::select(Bulk_VAF, id)

names(Cluster_check_filter_4_matched)[names(Cluster_check_filter_4_matched) == "Bulk_VAF"]<-"C4_VAF"

```

```{r}

Cluster_check_5_matched <- detect_mutations(C5_final)

Cluster_check_filter_5_matched <- Cluster_check_5_matched %>%
                                  dplyr::filter(GT_call_rate > 10) %>%
                                  dplyr::select(Bulk_VAF, id)

names(Cluster_check_filter_5_matched)[names(Cluster_check_filter_5_matched) == "Bulk_VAF"]<-"C5_VAF"

```

Then we'll merge annotated variant dataframes for all clusters, and search for mutations that were detected by bulk NGS at any timepoint (diagnosis, remission, and relapse (if applicable)). 
```{r}

#select id and amino acid columns from final_mutation_info
ref_mut<- final_mutation_info %>% dplyr::select(c("id","AA"))

clusters_for_hm_matched<- merge(ref_mut,Cluster_check_filter_1_matched, by = "id", all = T ) %>%
                          merge(., Cluster_check_filter_2_matched, by = "id", all = T) %>%
                          merge(., Cluster_check_filter_3_matched, by = "id", all = T) %>%
                          merge(., Cluster_check_filter_4_matched, by = "id", all = T) %>%
                          merge(., Cluster_check_filter_5_matched, by = "id", all = T) %>%
                          dplyr::filter(AA == "DNMT3A.R882H" | 
                                        AA == "NRAS.G13D" | 
                                        AA == "NRAS.G12D" | 
                                        AA == "IDH2.R140Q"|
                                        AA == "NPM1.287"| 
                                        AA == "KRAS.Q61P"| 
                                        AA == "RUNX1.118" | 
                                        AA == "IDH2.R172K"| 
                                        AA == "SRSF2.P95H"| 
                                        AA == "JAK2.V617F"| 
                                        AA == "DNMT3A.R736C"| 
                                        AA == "NRAS.G12V")%>%
                          column_to_rownames(., var = "AA") %>%
                          dplyr::select(!c("id"))
                  
clusters_for_hm_matched[is.na(clusters_for_hm_matched)]<- 0
names(clusters_for_hm_matched)[names(clusters_for_hm_matched) == "C1_VAF"]<- "Cluster 1"
names(clusters_for_hm_matched)[names(clusters_for_hm_matched) == "C2_VAF"]<- "Cluster 2"
names(clusters_for_hm_matched)[names(clusters_for_hm_matched) == "C3_VAF"]<- "Cluster 3"
names(clusters_for_hm_matched)[names(clusters_for_hm_matched) == "C4_VAF"]<- "Cluster 4"
names(clusters_for_hm_matched)[names(clusters_for_hm_matched) == "C5_VAF"]<- "Cluster 5"

clusters_for_hm_matched

```

Take a look at which mutations were present in which clusters. 
```{r}

pheatmap(as.matrix(clusters_for_hm_matched), 
         cluster_rows = TRUE, 
         cluster_cols = FALSE,
         color = inferno(8),
         breaks = c(0, 0.5, 1, 10, 20, 30, 40,50, 60),
         display_numbers = TRUE, 
         treeheight_row = 0,
         fontsize_number = 10, 
         number_color = "white", 
         fontsize_col = 10, 
         fontsize_row = 10)

```

Tally number of cells classified in each cluster
```{r}

ncells_C1 <- C1_final %>% nrow() %>% as.data.frame() %>% rename("." = "Cluster_1")
ncells_C2 <- C2_final %>% nrow() %>% as.data.frame() %>% rename("." = "Cluster_2")
ncells_C3 <- C3_final %>% nrow() %>% as.data.frame() %>% rename("." = "Cluster_3")
ncells_C4 <- C4_final %>% nrow() %>% as.data.frame() %>% rename("." = "Cluster_4")
ncells_C5 <- C5_final %>% nrow() %>% as.data.frame() %>% rename("." = "Cluster_5")

ncells <- cbind(ncells_C1, ncells_C2) %>%
          cbind(., ncells_C3) %>%
          cbind(., ncells_C4) %>%
          cbind(., ncells_C5) %>% reshape2::melt(.)

ncells

```

Calculate the proportion of cells successfully classified 
```{r}

percent_recovery <- sum(ncells$value)/nrow(filtered_NGT)*100

percent_recovery

```

This table which contains mutation, sample, and VAF info is exported, annotated in excel, and used for plotting in the "scMRD_decon_plots" script. 
```{r}

clusters_for_decon <- clusters_for_hm_matched %>% rownames_to_column(., var = "Mutations") %>%
                       reshape2::melt(.) %>% 
                       mutate(variable = str_replace(variable, "Cluster", "Sample")) %>% 
                       rename(variable = "Sample") %>% rename(value = "VAF")

head(clusters_for_decon)

```

Subset the filtered NGT based on the cells in each cluster and store as individual sample.s 
```{r}
 
 # sample_1 <- subset(filtered_NGT, Cell %in% C1_final$Cell)
 # sample_2 <- subset(filtered_NGT, Cell %in% C2_final$Cell)
 # sample_3 <- subset(filtered_NGT, Cell %in% C3_final$Cell)
 # sample_4 <- subset(filtered_NGT, Cell %in% C4_final$Cell)
 # sample_5 <- subset(filtered_NGT, Cell %in% C5_final$Cell)

```


Data are exported as RDS files which are used for downstream clonal + protein analyses. Please note that the sample naming during the demulitplexing may differ from other scripts in this project. The naming system (Cluster 1-5) in this file was set by the K-means clustering algorithm, and the samples were renamed during subsequent processing/plotting according to how many cells were recovered (i.e. S1 has highest cell number, S5 has lowest). 
```{r}
   
  # write_rds(sample_1, file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/WX_MRD5_v2_S1.rds")
  # write_rds(sample_2, file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/WX_MRD5_v2_S2.rds")
  # write_rds(sample_3, file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/WX_MRD5_v2_S3.rds")
  # write_rds(sample_4, file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/WX_MRD5_v2_S4.rds")
  # write_rds(sample_5, file = "/Users/robinst1/Desktop/MSKCC/Levine_Lab/Data/R/scDNA_analysis/WX/MRD_Manuscript/WX_MRD5_v2_S5.rds")

```


















